Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-12 Mag_quotients.tex
Relecture exemples Grassmann

We terminate this section with an example illustrating the lattice
constructions on $\QMag$ in the case where the
characteristic of the field $\K$ is different from $2$. For that, we
introduce various operads, which requires the following notations, also
used in Section~\ref{sec:CAs_d}. Let, for any integer $\gamma \geq 1$,
the binary trees $\LComb{\gamma}$ and $\RComb{\gamma}$ be respectively
the left and the right combs of degree $\gamma$. These trees are depicted
as
\begin{equation}
  \label{equ:left_and_right_combs}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal
nodes they contain.

We first recall that the \Def{linear associative operad} is
\begin{equation}
  \label{equ:linear_associative_operad}
  \KAs:=\KMag/_{\IAs}
\end{equation}
where $\IAs$ is the ideal spanned
\begin{math}
  \LComb{2}-\RComb{2},
\end{math}
and that its Hilbert series is
\Todo{référence ?}
\begin{equation}
    \label{equ:Hilbert_series_of_As}
    \HilbertSeries_{\KAs}(t)=\sum_{n\geq 1}t^n,
    \end{equation}

We define the \Def{anti-associative operad} by
\begin{equation}
  \label{equ:anti-associative_operad}
  \AAs:=\KMag/_{\IAAs}
\end{equation}
where $\IAAs$ is the ideal spanned by
\begin{math}
  \LComb{2}+\RComb{2}
\end{math}.
Using the Buchberger algorithm for operads~\cite[Section 3.7]{DK10}, we
check that the set of rewrite rules
\begin{equation}
  \label{equ:convergent_presentation_of_AAS}
  \left\{\LComb{2}\Rew -\RComb{2},\
  \RComb{3}\Rew 0
  \right\}
\end{equation}
is a convergent presentation of $\AAs$. We point out that this statement
is false if the characteristic of $\K$ is equal to $2$. Moreover, using
the convergent presentation~\eqref{equ:convergent_presentation_of_AAS},
we have
  \begin{equation}
    \label{equ:Hilbert_series_of_AAs}
    \HilbertSeries_{\AAs}(t)=t+t^2+t³.
    \end{equation}
\medbreak

We define the \Def{2-nihilpotente operad} by
\begin{equation}
  \label{equ:2-nihilpotente_operad}
  \NiHil:=\KMag/_{\INiHil}
  \end{equation}
where $\INiHil$ is the ideal spanned by the two trees $\LComb{2}$ and
$\RComb{2}$.
\Todo{Est-ce qu'il faut rajouter le 2 dans la macro désignant cette
  opérade ? De plus, Samuele,\\
  pourrais-tu rajouter la référence de Loday ?}
We have
\begin{equation}
  \label{equ:Hilbert_series_of_NiHil}
  \HilbertSeries_{\NiHil}(t)=t+t^2.
\end{equation}
\Todo{Est-ce bien la bonne série ? Samuele, tu avais l'air de dire que
  c'était plutôt $t+2t^2$. Ou peut-être que NiHil est la $1$-
  nihilipotente opérade ?}

Finally, for every integer $\gamma\geq 2$, we define the
\Def{$\gamma$-right comb operad} $\RC{\gamma}$ as follows. For every
arity $n$, we let
\begin{equation}
  \label{equ:compositions_of_RC}
  \RC{\gamma}(n):=
    \begin{cases}
        \KMag(n) &
        \mbox{if } n \leq \gamma, \\
        \K\Angle{\RComb{n}}
            & \mbox{otherwise},
    \end{cases}
\end{equation}
and the partial composition $\Tfr_1 \circ_i \Tfr_2$ is the partial
composition of $\Tfr_1$ and $\Tfr_2$ in $\Mag$ if the integer
\begin{math}
  n:=|\Tfr_1|+|\Tfr_2| 
\end{math}
is smaller than or equal to $\gamma$, and $\RComb{n}$, otherwise.
Moreover, by definition of the $\gamma$-right comb operad, we have
\begin{equation}
  \label{equ:Hilbert_series_of_RC}
  \HilbertSeries_{\RC{\gamma}}(t)=\sum_{1\leq n\leq\gamma}\Catalan(n) t^n
  +\sum_{n\geq\gamma+1}t^n.
  \end{equation}
\medbreak

\begin{Lemma}
  \label{lem:presentation_of_RC}
  We have an isomorphism
  \begin{equation}
  \label{equ:presentation_of_RC}
  \RC{\gamma}\simeq\KMag/_{\IRC{\gamma}}
  \end{equation}
  where $\IRC{\gamma}$ is the ideal spanned by the elements
  $\Tfr-\RComb{\gamma}$, where $\Tfr$ runs over all the binary trees of
  arity $\gamma+1$. In other words, $\RC{\gamma}$ is a combinatorial
  realization of
  \begin{math}
    \KMag/_{\IRC{\gamma}}.
    \end{math}
\end{Lemma}

\begin{proof}
  Let $R^{(\gamma)}$ be the set of rewrite rules
  $\Tfr\Rew\RComb{\gamma}$, where $\Tfr$ runs over all the binary trees
  of arity $\gamma+1$ different from $\RComb{\gamma}$. The unique normal
  form of arity $n\geq\gamma+1$ for  $R^{(\gamma)}$ is equal to
  $\RComb{n}$, so that $R^{(\gamma)}$ is a convergent presentation of
  \begin{math}
    \KMag/_{\IRC{\gamma}}.
  \end{math}
  Moreover, the normal forms of arity $n\leq\gamma$ are all the trees
  of arity $n$ and, by using the convergent presentation
  $R^{(\gamma)}$, the compositions of
  \begin{math}
    \KMag/_{\IRC{\gamma}}
  \end{math}
  satisfy~\eqref{equ:compositions_of_RC}. Hence, $R^{(\gamma)}$ is also a
  convergent presentation of
  \begin{math}
    \RC{\gamma}
  \end{math}
  which proves Lemma~\ref{lem:presentation_of_RC}.
\end{proof}

\medbreak
The lower-bound and the upper-bound of $\As$ and $\AAs$ in the lattice
$\LatQMag$ are given by the following.
\begin{Theorem}
  \label{thm:example_lattice}
  We have
  \begin{equation}
    \label{equ:example_lower-bound}
    \KAs\InfQMag\AAs=\NiHil
  \end{equation}
  and
  \begin{equation}
    \label{equ:example_upper-bound}
    \KAs\SupQMag\AAs=\RC{3}.
    \end{equation}
  \end{Theorem}

\begin{proof}
  First, we show~\eqref{equ:example_lower-bound}. The ideal of relations
  of $\KAs\InfQMag\AAs$ is equal to $\IAs+\IAAs$, so that it is spanned
  by the two elements $\LComb{2}-\RComb{2}$ and $\LComb{2}+\RComb{2}$. By
  linear transformations applied to these generators, $\IAs+\IAAs$ is
  spanned by $\LComb{2}$ and $\RComb{2}$, that is it is equal to
  $\INiHil$, which proves~\eqref{equ:example_lower-bound}.

  \smallbreak
  Now, we show~\eqref{equ:example_upper-bound}. We denote by
  \begin{math}
    \pi:\KMag\to\KAs\SupQMag\AAs
  \end{math}
  the natural projection. Let $\Tfr$ be a tree of arity $4$ and let
  us define $\alpha_{\Tfr}:=\Tfr-\RComb{3}$. The elements $\alpha_{\Tfr}$
  belong to $\IAs$ and to $\IAAs$ since both $[\Tfr]_{\IAAs}$ and to
  $[\RComb{3}]_{\IAAs}$ are equal to $[0]_{\IAAs}$. The last statement is
  shown using the convergent
  presentation~\eqref{equ:convergent_presentation_of_AAS}. Hence, the
  ideal generated by the elements $\alpha_{\Tfr}$ is included in 
  $\IAs\cap\IAAs=\ker(\pi)$, so that $\pi$ induces a surjective morphism
  \begin{math}
    \overline{\pi}:\RC{3}\to\KAs\SupQMag\AAs
  \end{math}
  from Lemma~\ref{lem:presentation_of_RC}. We conclude using Hilbert
  series: $\HilbertSeries_{\KAs\SupQMag\AAs}(t)$ is computed
  using the Grassmann formula analog with
  Formulas~\eqref{equ:Hilbert_series_of_As},
  \eqref{equ:Hilbert_series_of_AAs} and
  ~\eqref{equ:Hilbert_series_of_NiHil} and it turns out to be equal to
  $\HilbertSeries_{\RC{3}}(t)$ which is given
  in~\eqref{equ:Hilbert_series_of_RC}. Hence, $\overline{\pi}$ is an
  isomorphism, which proves~\eqref{equ:example_upper-bound}.
\end{proof}
\medbreak

>>>>

We terminate this section with an example illustrating the lattice
constructions on $\QMag$ in the case where the characteristic of the
field $\K$ is different from $2$. For that, we introduce various
operads, which requires the following notations, also used in
Section~\ref{sec:CAs_d}. Let, for any integer $\gamma \geq 1$, the
binary trees $\LComb{\gamma}$ and $\RComb{\gamma}$ be respectively the
left and the right combs of degree $\gamma$. These trees are depicted as
\begin{equation}
  \label{equ:left_and_right_combs}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal nodes
they contain.
\medbreak

We first recall that the \Def{linear associative operad} is
\begin{equation} \label{equ:linear_associative_operad}
    \KAs:=\KMag/_{\IAs}
\end{equation}
where $\IAs$ is the ideal spanned
\begin{math}
    \LComb{2}-\RComb{2},
\end{math}
and that its Hilbert series is
\begin{equation} \label{equ:Hilbert_series_of_As}
    \HilbertSeries_{\KAs}(t)=\sum_{n\geq 1}t^n,
\end{equation}
We define the \Def{anti-associative operad} by
\begin{equation}
    \label{equ:anti-associative_operad}
    \AAs:=\KMag/_{\IAAs}
\end{equation}
where $\IAAs$ is the ideal spanned by
\begin{math}
    \LComb{2}+\RComb{2}
\end{math}.
Using the Buchberger algorithm for operads~\cite[Section 3.7]{DK10}, we
check that the set of rewrite rules
\begin{equation}
    \label{equ:convergent_presentation_of_AAS}
    \left\{\LComb{2}\Rew -\RComb{2}, \; \RComb{3}\Rew 0 \right\}
\end{equation}
is a convergent presentation of $\AAs$. We point out that this statement
is false if the characteristic of $\K$ is equal to $2$. Moreover, using
the convergent presentation~\eqref{equ:convergent_presentation_of_AAS},
we have
\begin{equation} \label{equ:Hilbert_series_of_AAs}
    \HilbertSeries_{\AAs}(t)=t+t^2+t^3.
\end{equation}
Let us consider the \Def{2-nilpotent operad}~\cite{Zin12} defined by
\begin{equation} \label{equ:2-nilpotent_operad}
    \TwoNil:=\KMag/_{\INiHil}
\end{equation}
where $\INiHil$ is the ideal spanned by the two trees $\LComb{2}$ and
$\RComb{2}$. We have
\begin{equation} \label{equ:Hilbert_series_of_NiHil}
    \HilbertSeries_{\TwoNil}(t)=t+t^2.
\end{equation}
Finally, for every integer $\gamma\geq 2$, we define the
\Def{$\gamma$-right comb operad} $\RC{\gamma}$ as follows. For every
arity $n$, we let
\begin{equation} \label{equ:compositions_of_RC}
    \RC{\gamma}(n):=
    \begin{cases}
        \KMag(n) &
        \mbox{if } n \leq \gamma, \\
        \K\Angle{\RComb{n}} & \mbox{otherwise},
    \end{cases}
\end{equation}
\Todo{S: plutôt $\K\Angle{\RComb{n - 1}}$ pour avoir les bonnes arités,
non ?}
and the partial composition $\Tfr_1 \circ_i \Tfr_2$ is the partial
composition of $\Tfr_1$ and $\Tfr_2$ in $\Mag$ if the integer
\begin{math}
    n:=|\Tfr_1|+|\Tfr_2|
\end{math}
\Todo{S : je crois qu'il manque un "$-1$ ici, non ?}
is smaller than or equal to $\gamma$, is and $\RComb{n}$, otherwise.
Moreover, by definition of the $\gamma$-right comb operad, we have
\begin{equation} \label{equ:Hilbert_series_of_RC}
    \HilbertSeries_{\RC{\gamma}}(t)=
    \sum_{1\leq n\leq\gamma}\Catalan(n) t^n +\sum_{n\geq\gamma+1}t^n.
\end{equation}
\medbreak

\begin{Lemma} \label{lem:presentation_of_RC}
    We have an isomorphism
    \begin{equation} \label{equ:presentation_of_RC}
        \RC{\gamma}\simeq\KMag/_{\IRC{\gamma}}
    \end{equation}
    where $\IRC{\gamma}$ is the ideal spanned by the elements
    $\Tfr-\RComb{\gamma}$, where $\Tfr$ runs over all the binary trees
    of arity $\gamma+1$. In other words, $\RC{\gamma}$ is a
    combinatorial realization of
    \begin{math}
        \KMag/_{\IRC{\gamma}}.
    \end{math}
\end{Lemma}
\begin{proof}
    Let $\Rew$ be the set of rewrite rules $\Tfr\Rew\RComb{\gamma}$,
    where $\Tfr$ runs over all the binary trees of arity $\gamma+1$
    different from $\RComb{\gamma}$. The unique normal
    form of arity $n\geq\gamma+1$ for the rewrite relation $\RewContext$
    induced by $\Rew$ is $\RComb{n}$, so that $\Rew$ is a convergent
    presentation of
    \begin{math}
        \KMag/_{\IRC{\gamma}}.
    \end{math}
    \Todo{S: "$n - 1$" aussi ici ?}
    Moreover, the normal forms for $\RewContext$ of arity $n\leq\gamma$
    are all the trees of arity $n$ and, by using the convergent
    presentation $\Rew$, the compositions of
    \begin{math}
        \KMag/_{\IRC{\gamma}}
    \end{math}
    satisfy~\eqref{equ:compositions_of_RC}. Hence, $\Rew$ is also a
    convergent presentation of $\RC{\gamma}$ which proves the statement
    of the lemma.
\end{proof}
\medbreak
\Todo{S: définir d'abord dans le cadre ensembliste puis considérer son
analogue linéaire ?}

The lower-bound and the upper-bound of $\As$ and $\AAs$ in the lattice
$\LatQMag$ are described by the following.
\medbreak

\begin{Theorem} \label{thm:example_lattice}
    We have
    \begin{equation} \label{equ:example_lower-bound}
        \KAs\InfQMag\AAs=\TwoNil
    \end{equation}
    and
    \begin{equation} \label{equ:example_upper-bound}
        \KAs\SupQMag\AAs=\RC{3}.
    \end{equation}
\end{Theorem}
\begin{proof}
    The ideal of relations of $\KAs\InfQMag\AAs$ is equal to
    $\IAs+\IAAs$, so that it is spanned by the two elements
    $\LComb{2}-\RComb{2}$ and $\LComb{2}+\RComb{2}$. By linear
    transformations applied to these generators, $\IAs+\IAAs$ is spanned
    by $\LComb{2}$ and $\RComb{2}$, that is, it is equal to $\INiHil$,
    which proves~\eqref{equ:example_lower-bound}.
    \smallbreak

    Let us now denote by
    \begin{math}
        \pi:\KMag\to\KAs\SupQMag\AAs
    \end{math}
    the natural projection. Let $\Tfr$ be a tree of arity $4$ and let us
    define $\alpha_{\Tfr}:=\Tfr-\RComb{3}$. The elements
    $\alpha_{\Tfr}$ belong to $\IAs$ and to $\IAAs$ since both
    $[\Tfr]_{\IAAs}$ and to $[\RComb{3}]_{\IAAs}$ are equal to
    $[0]_{\IAAs}$. The last statement is shown by using the convergent
    presentation~\eqref{equ:convergent_presentation_of_AAS} of $\AAs$.
    Hence, the ideal generated by the elements $\alpha_{\Tfr}$ is
    included in $\IAs\cap\IAAs=\ker(\pi)$, so that $\pi$ induces a
    surjective morphism
    \begin{math}
        \bar{\pi}:\RC{3}\to\KAs\SupQMag\AAs
    \end{math}
    from Lemma~\ref{lem:presentation_of_RC}. We conclude by using
    Hilbert series: $\HilbertSeries_{\KAs\SupQMag\AAs}(t)$ is computed
    by using the Grassmann formula analog with
    Formulas~\eqref{equ:Hilbert_series_of_As},
    \eqref{equ:Hilbert_series_of_AAs},
    and~\eqref{equ:Hilbert_series_of_NiHil}, and it turns out to be
    equal to $\HilbertSeries_{\RC{3}}(t)$ which is given
    in~\eqref{equ:Hilbert_series_of_RC}. Hence, $\bar{\pi}$ is an
    isomorphism, which proves~\eqref{equ:example_upper-bound}.
\end{proof}
\medbreak
